---
title: "ADJ Final Project - CKD analysis"
author: "Joy Mazur"
date: "2024-04-27"
output: html_document
---

Query 1 downloaded morning 4/27/24. 
Underlying Cause of Death, 1999-2020
Queried ICD-10 codes N17-N19 (all having to do with renal failure, associated with chronic kidney disease)
URL: https://wonder.cdc.gov/controller/saved/D76/D389F388
NOTE: Total deaths are 979,364, total population 6,746,356,647, total crude rate 14.5	

Query 2 downloaded morning 4/27/24
Multiple Cause of Death, 1999-2020
Queried ICD-10 codes H40, I10-I15, I20-I25, E10-E14, N25.1, I51.6 (associated with diseases that are comorbidities of CKD)
URL: https://wonder.cdc.gov/controller/saved/D77/D389F387
NOTE: Total deaths are 20,065,194, total population 6,746,356,647, total crude rate 297.4

Loading libraries
```{r}
library(tidyverse)
library(janitor)
library(data.table)
library(lubridate)
```

Loading first file
```{r}
ckd_death <- read_delim("data/ckd_cdc.txt", delim = "\t", col_types ="ccniiin")

?col_types
```

Remove the notes + clean column names
```{r}
ckd_death %>% 
  filter(!is.na(Notes)) %>% 
  select(Notes)
  
ckd_death <- ckd_death %>% 
  filter(is.na(Notes)) %>% 
  select(-Notes)

ckd_death <- clean_names(ckd_death)
```

Playing around with the data
```{r}
ckd_death %>% 
  arrange(desc(deaths))
#Texas has a lot of recent years where they have the most deaths - not surprising since they have a huge population

ckd_death %>% 
  summarize(tot_death = sum(deaths, na.rm=T))
#Total deaths attributed to CKD from 1999-2020 is 979,364
#This lines up with the query results I received from the in-page table on the site

ckd_death %>% 
  group_by(state) %>% 
  arrange(desc(crude_rate))
#West Virginia had the highest crude rate in 2020 at 30.7

ckd_death %>% 
  filter(year=="1999") %>% 
  arrange(desc(crude_rate))
#Alabama had the highest crude rate: 22.0

ckd_death %>% 
  group_by(state, deaths) %>% 
  filter(year_code=="1999") %>% 
  arrange(desc(crude_rate))
#discovered that "2020" is not present in the year column but is present in the year_code column - using year_code column is probably safest here

ckd_death %>% 
  filter(year_code=="2020", state=="Alabama")
#Alabama's crude rate of death in 2020 was 21.8, just a little lower than it was in 1999. This shows not much progress has been made in that state at least

ckd_death %>% 
  group_by(state, deaths) %>% 
  filter(year_code=="2020") %>% 
  arrange(desc(crude_rate))

mo_ckdrate <- ckd_death %>% 
  group_by(state, deaths) %>% 
  filter(state=="Missouri") %>% 
  arrange(desc(crude_rate))
#While not exactly linear, Missouri's crude rate of CKD death has only really increased over time
#Highest crude rate of death was 2020 with a crude rate of 27.0 

ggplot(mo_ckdrate, aes(x=year_code, y=crude_rate)) +
  geom_line() +
  labs(title="Crude rate of CKD deaths in Missouri over time", 
       y="Crude Rate",
       x="Year")
#Has increased over time

ckd_death %>% 
  filter(state=="Missouri") %>% 
  summarize(tot_death=sum(deaths, na.rm=T))
#total deaths = 26,986 deaths

ckd_death %>% 
  filter(state=="Missouri") %>% 
  summarize(tot_pop=sum(population,na.rm=T))
#total population = 130,170,885 people

(26986/130170885)*100000
#Missouri crude rate over time = 20.7 
```

Creating new dataframe for rate over time
```{r}
#THIS IS THE RIGHT WAY
ckd_state_totrate <- ckd_death %>% 
  group_by(state) %>% 
  summarize(tot_death=sum(deaths, na.rm=T), tot_pop=sum(population, na.rm=T)) %>% 
  mutate(tot_rate = tot_death/tot_pop*100000) %>% 
  arrange(desc(tot_rate))
#West Virginia had the highest overall rate of death with 24.3. Missouri ranked eighth


ckd_year_totrate <- ckd_death %>% 
  group_by(year_code) %>% 
  summarize(tot_death=sum(deaths, na.rm=T), tot_pop=sum(population, na.rm=T)) %>% 
  mutate(tot_rate = tot_death/tot_pop*100000) %>% 
  arrange(desc(tot_rate))
#2020 had the highest national CKD death rate with a crude rate of 15.6

ggplot(ckd_year_totrate, aes(x=year_code, y=tot_rate)) +
  geom_line() +
  labs(title="National crude rate of CKD deaths over time", 
       x="Year", 
       y="National crude rate")

#ckd_totrate %>% 
  #summarize(tot_rate= sum(tot_death/tot_pop)*100000)
#um no this is wrong - need to go back to comments from Liz to figure out 
#UPDATE: FIXED AND THE TOP ONE IS RIGHT


#LIZ NOTES: 
opioid_deaths %>% 
  group_by(year) %>% 
  summarise(tot_deaths = sum(deaths, na.rm=T), tot_pop = sum(population, na.rm=T)) %>% 
  mutate(yearly_rate = tot_deaths/tot_pop*100000) %>% 
  arrange(desc(yearly_rate))
```


Load second file 
```{r}
comorb <- read_delim("data/comorb.txt", delim = "\t", col_types="ccniiiin")
```

Remove notes + clean column names
```{r}
comorb %>% 
  filter(!is.na(Notes)) %>% 
  select(Notes)
  
comorb <- comorb %>% 
  filter(is.na(Notes)) %>% 
  select(-Notes)

comorb <- clean_names(comorb)
```



Going through some of the same questions/playing around with this dataset
```{r}
comorb %>% 
  summarize(tot_death = sum(deaths, na.rm=T))
#Total amount of deaths is 20,065,194 
#Matches up with the results of the query on the webpage 

comorb %>% 
  arrange(desc(deaths))
#California in 2020 had the most deaths out of anything else - again, not surprising because of population there

comorb %>% 
  filter(year_code=="1999") %>% 
  arrange(desc(crude_rate))
#West Virginia had the highest crude rate with 468.5 deaths per 100,000

comorb %>% 
  filter(year_code=="2020") %>% 
  arrange(desc(crude_rate))
#Oklahoma had the highest crude rate with 641.0 deaths per 100,000 

comorb %>% 
  filter(year_code=="2020") %>% 
  arrange(desc(population))
#California in 2020 had the highest population

mo_comrate <- comorb %>% 
  group_by(state, deaths) %>% 
  filter(state=="Missouri") %>% 
  arrange(desc(crude_rate))
#Missouri had the highest rate of death in 1999, with a crude rate of 380.4 
#Again not linear, but it seems like the rate generally decreased over time vs. CKD rate increasing over time. That is interesting

ggplot(mo_comrate, aes(x=year_code, y=crude_rate)) +
  geom_line() +
  labs(title="Crude rate of CKD comorbidity deaths in Missouri over time", 
       y="Crude Rate",
       x="Year")
#Mostly decreased, with a big increase in comorbidity in 2020
  
```

Creating new dataframe for rates over time for the comorb dataset, by state
```{r}
comorb_state_totrate <- comorb %>% 
  group_by(state) %>% 
  summarize(tot_death=sum(deaths, na.rm=T), tot_pop=sum(population, na.rm=T)) %>% 
  mutate(tot_rate = tot_death/tot_pop*100000) %>% 
  arrange(desc(tot_rate))
#West Virginia had the highest total death rate from comorbidities over time with a rate of 455.3 
```

Creating new dataframe for national crude rates by year
```{r}
comorb_year_totrate <- comorb %>% 
  group_by(year_code) %>% 
  summarize(tot_death=sum(deaths, na.rm=T), tot_pop=sum(population, na.rm=T)) %>% 
  mutate(tot_rate = tot_death/tot_pop*100000) %>% 
  arrange(desc(tot_rate))
#2020 has the highest crude rate of death with a rate of about 367

ggplot(comorb_year_totrate, aes(x=year_code, y=tot_rate)) +
  geom_line() +
  labs(title="National crude rate of CKD comorbidity deaths over time", 
       x="Year", 
       y="National crude rate")
```


Questions I want to ask of my data: 

What is the difference between deaths attributed to CKD and deaths attributed to JUST any comorbidities of CKD? 
```{r}
#CKD: 979,364
#comorb: 20,065,194 
#ANS: A difference of 19,085,830. The CKD deaths are 4% of the comorbidity deaths. How many of these comorbidity deaths could have an underlying cause of chronic kidney disease? 
```

What states have the highest rates of deaths attributed to CKD? Is this the same as the states that have the highest rates of death attributed to comorbidities of CKD? 

#Top 5 for CKD: 
1. West Virginia (24.3)
2. Louisiana (23.6)
3. Mississippi (23.2)
4. Pennsylvania (22.6)
5. Arkansas (22) 

8. Missouri (20.7)

#Top 5 for comorbidities: 
1. West Virginia (455.3)
2. Oklahoma (429.2)
3. Rhode Island (390.1)
4. Ohio (375.2)
5. Mississippi (369.2)

18. Missouri (325.9)



What year had the highest crude rate over time for CKD and for comorbidities?

#For CKD: 
2020, rate 15.6
#For comorbidities: 
2020, with rate about 367.0

How has the rate of CKD and comorbidities fared over time nationally? For Missouri? 
```{r}
#CKD in Missouri: Crude rate of CKD deaths has generally increased over the past two decades
#CKD nationally: Generally increased over time
#Comorbidities in Missouri: Generally decreased, with a spiked increase in 2020
#Comorbidities nationally: Generally decreased until about 2015 when it rose substantially, especially starting in about 2018. 
```


Overall, what questions would this prompt for further reporting on this issue? Can this data actually help us make any educated guesstimations about the apparent underreporting of chronic kidney disease as a cause of death? 


^^Answer these questions, come up with two more. Find some way to join the tables? Make a ggplot? Stacked bar graph ? - crude rate CKD vs. crude rate comorb

```{r}
cbind(comorb_year_totrate, ckd_year_totrate)

#figured out I'd have to rename columns in the original tables here to do this... this is probably not the easiest way to do this but whatever 
  
?colnames()

rate_join <- inner_join(ckd_year_totrate, comorb_year_totrate, by = c("year_code"="year_code")) %>% 
  select(-tot_death.x, -tot_pop.x, -tot_death.y, -tot_pop.y) %>% 
  mutate(ckd_rate = tot_rate.x, comorb_rate = tot_rate.y) %>% 
  select(-tot_rate.x, -tot_rate.y)
  #I don't know why this isn't working
  #UPDATE - it was a syntax issue
#YES THIS IS WHAT I WANTED SLAY

rate_join %>% 
pivot_longer()
ggplot(rate_join, aes(fill=comorb_rate, ckd_rate, x=year_code, y=comorb_rate)) + 
         geom_bar(position="dodge", stat="identity")
#need to figure out how to do this - pretty much the last thing I need to do? 
 
longrate_join <- rate_join %>%
  pivot_longer(ckd_rate:comorb_rate)

ggplot(longrate_join, aes(x=year_code, y=value, fill=name)) +
  geom_bar(position="dodge", stat="identity") +
  labs(title="CKD vs. CKD comorbidity death rates",
       x="Year",
       y="Rate",
       fill="Disease")+
  scale_fill_discrete(labels=c("CKD", "Comorbidity"))

#YES I WANT THIS, NOW JUST CLEAN IT W/LABELS

```



Any problems or limitations with the datasets?
- All entries for 2020 have 2020 in the year_code column but not in the year column. Because of this, when grouping or filtering by year, I always used the year_code column for accuracy